\bf Model Visualization Techniques for a Social Network Model

### 2018-09-26 paper-child.Rnw
done with sec 4 edits, onto sec 3 edits. sec 2,1,A,6 still to go.

Every good data analysis includes both numerical and visual summaries of the data, so why restrict model description and diagnostics to numerical summaries? The concept of \emph{model visualization} was developed to complement traditional model diagnostic tools such as $R^2$ values and residual plots \citep{modelvis}. In a modeling task, there are three different things called ``model" under consideration: the model \emph{family}, the model \emph{form}, and the \emph{fitted} model. The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters. 
The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family. % In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. 
Finally, the \textit{model family} is the broadest description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand. %For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.

The model family, the model form, and the fitted model can each be visualized according to the three principles of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result. Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form. Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox. %We begin by introducing the five models that we fit to our example data. 
We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization. %We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.

@@

Now that we have assessed model fit, we next want to learn about model behavior in order to choose better fitting models in the future. We examine the behavior of SAOMs by visualizing collections of models. We build collections of models in four ways. First, we explore the space of all possible models to determine the most significant model parameters.  Second, we vary model settings to uncover relationships between parameters and their effects on the fitted model. Next, we fit the same model to the same data many times to assess the behavior of the underlying simulations. Finally, we fit the same model to different data to learn about fitted parameter values under different conditions. We chose these four collections because they each explore something different about SAOM behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances. Using the tools in our visualization toolbox, we %The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary. The third collection shows how those many possible parameters affect the values of parameters in the fitted models. Finally, we explore how the same model behaves under two completely different circumstances.

@@

\noindent \textbf{Fitting the same model to different data:} The final procedure for viewing collections of models is to fit the same model to different data to understand how the model behaves in different contexts. As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data 100 times each. The means and standard deviations of the parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix. Looking at the estimates for both data sets, we can see a few patterns emerging. First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$. In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five. %In addition, the outegree parameter, $\beta_1$, \textit{decreases} when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases. This inverse relationship implies that the girls prefer to form small, closely knit friend groups, as opposed to being popular and having many friends. Tight-knit friend groups are more encouraged by model M2 than by models M1 and M3.
This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not. When we Consider the context, this suddenly seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills. We are also concerned, however, with the extremely high correlation between the estimates of $\beta_1, \beta_2$. In all model-data combinations, the absolute correlation between the estimates is above 0.93. %We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
% We have seen how the same model forms behave in different context, with some  differences and some similarities. 
XXX NEED CONCLUDING SENTENCES XXX

We saw earlier that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model. We see in Figure~\ref{fig:compareModels} that this effect is not present in the senate data. Furthermore, the estimates of $\beta_4$ for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not informative for the senate data. But in the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are strongly discouraged from forming, probably because teenage girls want to have direct friendships with each other. Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial at times. %Although this indirect relationship variable does not help explain the senate collaboration structure, it does help clarify the teenage friendship structure.

@@

When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user. One of the hidden steps in the SIENA method of moments algorithm is the \textit{microstep} process. At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of differences between two network observations. We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain. Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions. But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process. We use these steps to visually explore the algorithms, so that we are visualizing the process, and not just the main result.

>>>>

%Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries? 
\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots by enriching our understanding and interpretation of statistical models. Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg. 2]{modelvis}, can benefit from model visualization (model-vis). We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}. Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form. Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}. %In a modeling task, there are three different things called ``model" under consideration: . The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters. 
%The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family. % In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. 
%Finally, the \textit{model family} is the broadest description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand. %For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
%The model family, the model form, and the fitted model can each be visualized according to  %We begin by introducing the five models that we fit to our example data. 
%We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization. %We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.

@@

Now that we have assessed model fit, we next want to learn about model behavior in order to choose better fitting models in the future. This is the second principle of model visualization: "collections [of models] are more informative" \citep{modelvis}. We examine the behavior of SAOMs by visualizing collections of models. We build collections of models in four ways. First, we explore the space of all possible models to determine the most significant model parameters.  Second, we vary model settings to uncover relationships between parameters and their effects on the fitted model. Next, we fit the same model to the same data many times to assess the behavior of the underlying simulations. Finally, we fit the same model to different data to learn about fitted parameter values under different conditions. We chose these four collections because they each explore something different about SAOM behavior: the choice of parameters in the objective function, the effects of underlying simulations, and the resulting fitted values under different circumstances. Using the tools in our visualization toolbox, we %The first considers the many parameters to include in a SAOM, while the second results in different parameter estimates, so we can see how parameter estimates vary. The third collection shows how those many possible parameters affect the values of parameters in the fitted models. Finally, we explore how the same model behaves under two completely different circumstances.

@@

\noindent \textbf{Fitting the same model to different data:} Finally, we view collections of models by fitting the same models to different data to understand how the model behaves in different contexts. As we did for the friendship data, we estimated the parameters for M1, M2, and M3, when fit to the senate data (see Figure \ref{fig:senateplot}) 100 times each. The means and standard deviations of the repeated parameter estimates for all six model-data combinations are given in Table~\ref{tab:meanests}, while corresponding density plots of the objective function parameters are given in Figure~\ref{fig:compareModels} in the Appendix. Looking at the estimates for both data sets, a few patterns emerge. First, we can see the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$ in both data sets and across all three models. The estimates of $\beta_1$ are all negative and range from about -5 and -3, while the estimates of $\beta_2$ are all positive and range from about four to five. %In addition, the outegree parameter, $\beta_1$, \textit{decreases} when $\beta_3$ is included, while the reciprocity parameter, $\beta_2$ increases. This inverse relationship implies that the girls prefer to form small, closely knit friend groups, as opposed to being popular and having many friends. Tight-knit friend groups are more encouraged by model M2 than by models M1 and M3.
This suggests that in both data sets, nodes prefer outgoing ties that are reciprocated to those that are not. When we consider the context, this seems obvious: the students want to have close mutual friendships, and senators want mutual support for their bills. We are also concerned, however, with the extremely high correlation between the estimates of $\beta_1, \beta_2$. In all six model-data combinations, the absolute correlation between $\hat{\beta}_1, \hat{\beta}_2$ is greater than 0.93. %We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.
% We have seen how the same model forms behave in different context, with some  differences and some similarities. 
After seeing the high correlation between parameter estimates in very different data contexts, we question the guideline that both parameters need to be included in the objective function.

In Figure~\ref{fig:distests}, we saw that the inclusion of $\beta_3$ for the friendship data has a noticeable effect on the other parameters in the model. In Figure~\ref{fig:compareModels}, however, we see that this effect is not present in the senate data. Furthermore, the estimates of $\beta_4$ for the senate data are near zero, suggesting that this effect, which considers indirect ties, is not informative for the senate data. In contrast for the friendship data, the mean estimate of $\beta_4$ is about -7.5, which suggests that indirect ties are \textit{strongly} discouraged from forming. Again, this is fairly intuitive given the context: teenage girls likely want to have direct friendships with each other. Senators, on the other hand, have to wheel and deal to pass their legislation, and indirect relationships may be beneficial at times. By examining the results of models fitted to different data, we better understand the interpretation of the objective function parameters, and we have learned of some potential parameter correlation problems affecting the fit of the models. All in all, we have used four different ways of viewing collections of models, which have given us insight into the behavior of SAOMs.

%Although this indirect relationship variable does not help explain the senate collaboration structure, it does help clarify the teenage friendship structure.

@@

The last model visualization principle, "explore the process of model fitting," is crucial for understanding SAOMs \citep{modelvis}. When fitting a SAOM in \texttt{RSiena}, several simulation steps are hidden from the user. One of the hidden steps in the SIENA method of moments algorithm is the \textit{microstep} process. At each microstep, an \textit{ego} node is selected to make a change, and the chosen ego node randomly makes one change in its ties to maximize its objective function. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of differences between two network observations. We want to simulate and view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain. Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions. But in order to understand \textit{how} the models are fit, we need to extract and explore the different steps of the process. We use these steps to visually explore the algorithms, so that we are visualizing the process, and not just the main result.
